---
title: "Phylogenetic tree of sea urchins"
author: "Emilia Jarochowska"
format: pdf
---

## Preparation

1. Create a new project in R Studio in a dedicated folder for this practical, using File -> New Project...

2. In the folder where your project is located, create a sub-folder `data`

3. Save the character matrix as `character_matrix.csv` in the `data` folder

4. Install the packages `phangorn` and `ape`. You will be likely asked to install some additional packages ("dependencies") - choose "yes".

5. Create a new script using File -> New File -> R Script, where you will put all the steps of the analysis. Save it in the project folder.

## Analysis

```{r}
library(phangorn)
library(ape)
```

Read your character matrix into R and convert it into the Nexus format.

*If you created your file using Excel on a system with Dutch language settings, your `csv` file likely uses a colon (;) instead of a comma (,) to separate columns. That is because in Dutch comma is used as a decimal separator, so to separate columns you need a different character. That is why the option `sep=";"` is used in this script. If your file uses a different character to separate columns, change the code accordingly.*

```{r}
Character_matrix <- read.csv(file="data/character_matrix.csv", 
                             header=T, 
                             sep=";", 
                             colClasses=rep("character", 8), # if you have 8 characters
                             row.names = 1) 
```

Convert the table into a character matrix format understood by `phangorn`:

```{r}
mm_pd <- phangorn::phyDat(as.matrix(Character_matrix), 
                          type = "USER", 
                          levels = 0:8)
```

Generate a distance matrix using the Hamming distance and examine the output:

```{r}
dm = phangorn::dist.hamming(mm_pd)
dm
```

Can you read from the distance matrix:

1. Which two taxa are the closest to each other?

2. Which ones are equally distant?

From the distance matrix, create a starting tree using the Neighbor-Joining method:

```{r}
start_tree <- ape::nj(dm)
plot(start_tree)
```

For unordered morphometric data, the option `model="ER+ASC"` should be used.

- ER: equal rates, all transitions occur at the same rate

- ASC: ascertainment bias correction (Lewis 2001) corrects for the fact that there are no constant sites (as is mostly the case with morphometric data); otherwise the branch lengths may be overestimated

```{r}
ml_nni <- phangorn::pml_bb(mm_pd,
             start = start_tree,
             model="ER+ASC", 
             rearrangement = "NNI")
plot(ml_nni,
     main = "NNI estimation")
```

Is this tree good? Can you see any polytomies?

Display a summary of the tree estimate and verify that the right model has been used:

```{r}
ml_nni
```

```{r}
rooted_ml_tree <- root(tree_ml, "Collohmannia_johnstoni", resolve.root = TRUE,
                             edgelabel=TRUE)
rooted_ml_tree <- pretty_root(rooted_ml_tree)
```