---
title: "Global plant diversity"
author: "Rike Wagner-Cremer, Iris Kuipers and Emilia Jarochowska"
format: docx
---

```{r setup}
#| include: false
library(paleobioDB)
library(dplyr)
```

## Before you start

Make a new folder for files related to this practical. In R Studio, choose `File` and `New project...`. Place the project in the existing folder which you have just created. Whenever you re-open this project, the paths to files will be automatically set.

## Objectives

After this practical, you will be able to:

*paleobiology skills*
- Create a diversty (richness) plot for a given taxon
- Explain the limitations of fossil data in calculating past diversity
- Assess how factors such as sampling effort, taxonomic and temporal resolution affect diversity analyses
*data skills*
- read in functions and apply them to new datasets
- present several variables in one plot

## Downloading data from the Paleobiology Database

You can download datasets using the web interface or by sending a direct query to the server using the database's API. The latter you can do using a script, this way you don't need to remember what to click. There are ready-made functions which make it easy to download data directly from your script. In the first example, you will use the function `pbdb_occurrences` from the package [`paleobioDB`](https://docs.ropensci.org/paleobioDB/index.html) by Varela et al. (2015). You need to install this package first.

### Analysing Pteridophyta

In this practical, you will analyse [Pteridophyta](https://en.wikipedia.org/wiki/Pteridophyte), [Pinophyta](https://en.wikipedia.org/wiki/Conifer) and [Magnoliophyta](https://en.wikipedia.org/wiki/Flowering_plant). The examples below will guide you through the analysis for Pteridophyta. Afterwards, you will do the same analysis for the other two divisions.

```{r download the occurrences}
Pteridophyta <- pbdb_occurrences(
  base_name = "Pteridophyta",
  show = c("coords", "classext"),
  vocab = "pbdb",
  limit = "all"
)
```

## 1. Genus richness through time	

Make a graph showing the genus richness of Pteridophyta over time.

```{r genus richness automatic}
#| fig-cap: "Genus richness of Pteridophyta"
pbdb_richness(Pteridophyta, 
              rank = "species", 
              temporal_extent = c(0, 540), 
              res = 20)
```
What happens if you change the parameter `res` (temporal resolution)? How do you decide which resolution is the best?

### Make the same plot but for all three groups: Pteridophyta, Pinophyta and Magnoliophyta.

But the plot made by the `paleobioDB` package doesn't allow plotting several groups on the same plot. What now? You can exploit the fact that the function `pbdb_richness` returns not only a plot, but a dataframe of richness per bin. You can save this dataframe into an object, here called `Pteri_rich`, and use it to add the remaining two groups to the plot using standard R plotting tools.

```{r extract richness table}
Pteri_rich <- pbdb_richness(Pteridophyta, 
              rank = "genus", 
              temporal_extent = c(0, 540), 
              res = 20,
              do_plot = F)
```

```{r plot richness from a table}
barplot(names.arg = Pteri_rich$temporal_intervals,
     height = Pteri_rich$richness)
```

Try to add the other groups on your own. You can use the parameter `add` of the `barplot` function or make a grouped barplot.

## 2. How to deal with records at different taxonomic levels?

A lot of the time fossils cannot be identifed to the species level. Also the concept of species in some groups is not clear. So many analyses rely on the genus level.

Look into the dataframe with the occurrences. Some are observations at the species level, some at the genus level. If you want to analyze richness at the genus level, you have to *somehow* include also those occurrences which had been made at the species level and discard the species name. E.g. one of the occurrences is *Metaclepsydropsis duplex*. You still want to count it, but as *Metaclepsydropsis*.
And you still have to consider that some occurrences are of different taxonomic ranks!

This code will show you what taxonomic ranks are present in the dataset:

```{r convert taxonomic ranks to a factor}
Pteridophyta$accepted_rank <- as.factor(Pteridophyta$accepted_rank)
levels(Pteridophyta$accepted_rank)
```

For the rest of the practical you only need records that are *at least* at the genus level (so species, subgenus and subspecies are ok). You can remove the remaining occurrences. Observe how the dimensions of the dataframe change.

```{r select only genus and below levels}
Pteridophyta <- Pteridophyta[Pteridophyta$accepted_rank %in% c("genus", "species", "subspecies","subgenus"),]
```

Now you can extract genus names from those names which also contain species and subspecies. You will need this again so it will be handy to have a function for it. This function takes all rows where the rank is different than `genus` and extracts the first word of the name. 

```{r function to extract genera from species}
species_to_genera <- function(df) {
  df <- df %>%
  mutate(
    genus_name = ifelse(
      accepted_rank != "genus",
      gsub(" .*", "", accepted_name),
      accepted_name
    )
  )
}
```

Now that the function is in memory, you can use it by applying it to the dataset:

```{r extrat genera from Pteridophyta}
Pteridophyta <- species_to_genera(Pteridophyta)
```

So now your analysis will use all the available records which *include* information about the genus.

## 3. How precisely do we know the age of a fossil occurrence?

To understand the nature of occurrence data, check in the dataframe how age information is provided. Which variables contain it? What is the average, minimal and maximal time span accross the occurrences? In other words, how *precise* is the age information for each occurrence?

```{r function for calculating duration}
calculate_duration <- function(df, max_ma, min_ma, occ_duration) {
  df[[occ_duration]] <- df[[max_ma]] - df[[min_ma]]
  return(df)
}
```

Apply the function to your dataset:

```{r calculate occurrence duration for Pteridophyta}
Pteridophyta <- calculate_duration(Pteridophyta, "max_ma", "min_ma", "occ_duration")
```

The last column in the dataframe, `occ_duration` is now the time span assigned to the occurrence, in My.

```{r distribution of occurrence durations}
avg_span <- mean(Pteridophyta$occ_duration)
hist(Pteridophyta$occ_duration,
     xlab = "Occurrence time span [My]",
     main = "Distribution of time spans of pteridophyte occurrences")
abline(v = avg_span,
       col = "red")
```

What is the average time span of an occurrence in the dataset? Does it mean the fossil lived for so long? Why are some durations so extremely long, i.e. having very low precision? If you were to cull the dataset to eliminate the most imprecisely dated occurrences, what precision would you accept as good? 

## 4. How is richness calculated?

In each consecutive time bin, a given taxon may be observed or not. If there are not many outcrops of rocks of given age on Earth, this age may yield no occurrences even though the organism existed. 

Check how the occurrences are distributed in time for a genus of your choice.

You will enocounter the problem: what time should you assign the occurrence to? You could plot it across the entire duration in which the given occurrence is recorded or use the mid-point (between minimum and maximum age) as the time coordinate.

```{r select a genus to plot}
selected_genus <- "Deltoidospora"
Deltoidospora_df <- Pteridophyta[Pteridophyta$genus == selected_genus, ]
```

Function for plotting age ranges:

```{r function for plotting age ranges}
plot_age_ranges <- function(df) {
  if (!all(c("min_ma", "max_ma") %in% colnames(df))) {
    stop("The dataframe must contain 'min_ma' and 'max_ma' columns.")
  }
  
  colors <- colorRampPalette(c("red", "blue"))(nrow(df))
  
  plot(
    x = NA, y = NA,
    ylim = c(1, nrow(df)), 
    xlim = c(min(df$min_ma), max(df$max_ma)), 
    ylab = "Occurrence", 
    xlab = "Age [Ma]",
    main = "Distribution of occurrences",
    type = "n" 
  )
  
  for (i in 1:nrow(df)) {
    lines(
      y = c(i, i), 
      x = c(df$min_ma[i], df$max_ma[i]), 
      lwd = 2 ,
      col = colors[i]
    )
  }
}
```

Apply this function to the genus you selected:

```{r plot age ranges for a selected genus}
plot_age_ranges(Deltoidospora_df)
```

What can you see on this plot? Are the occurrences well resolved in terms of time or poorly? Are there any occurrences with suspicious ages, e.g. when pteridophytes didn't exist?
Are there any times when this genus had not been found, but existed before and after? Is this possible? What could cause such distribution? 

**Make such plots for four more genera, choosing two that are very common and two that are rare.**

### 4.a. Dealing with temporal resolution 

Genus richness is calculating by summing occurrences recorded in a given time bin. But now you have seen that some occurrences are fond in a very wide time bins and some in narrow ones. So how do you compare occurrences with imprecise age information and with a precise one?

The answer is: [interpolation](https://en.wikipedia.org/wiki/Interpolation). Look at your first plot. The horizontal axis is divided into equal intervals and when you change the temporal resolution, the richness is recalculated to the new resolution. So you can request a plot with a temporal resolution of 2 My, but now you have seen that **most of your occurrence data does not have such a resolution**. So if you don't keep that in mind, you can create a plot that looks very precise, but it is based on imprecise data.

In interpolation, the time axis is divided into even intervals and the *density* of occurrences calculated from the durations of each occurrence is sampled so that you have one value of richness at each interval. 

So what is an honest temporal resolution at which you should analyse richness? Go back to chapter 3 above and find the **average duration** of an occurrence in your dataset. This is a reasonable candidate for temporal resultion of your analysis. Is it higher or lower than what you have used in your first plot?

### 4.b. Dealing with sampling issues

In the example where you saw occurrence durations of *Deltoidospora*, there were intervals when it had not been recorded. But of course it must have existed on Earth. There is a difference between genera that have been entered into the database and genera that are not in the database at that time, but should have been present at that time. There are terms specific to these two situations:

1. **sampled in bin**: only occurrences that are really recorded in this interval in the database
2. **range-through**: occurrences (and stratigraphic ranges) where the taxon is *inferred* to have existed even though it had not been found or entered into the database so far.

Range-through richness analysis assumes that a taxon is present in all bins between its first and last occurrence in the database.

Make a plot of richness of *Deltoidospora* and compare it with the plot you made for the duration of its occurrences.

```{r genus richness for a selected genus}
Deltoidospora_rich <- pbdb_richness(Deltoidospora_df, 
              rank = "genus", 
              temporal_extent = c(0, 540), 
              res = 20,
              do_plot = T)
```

Is this plot showing sampled in bin richness or range-through richness?

## 5. Sampled in bin richness

You could calculate sampled from the dataset you downloaded, but it would involve a lot of coding. The Paleobiology Database has tools to perform common calculations for you.
In the Download page you can select the type of data you want to download. To get diversity data for Pteridophyta at the genus level and the temporal resolution of a geochronological age, use [this link](https://paleobiodb.org/data1.2/occs/diversity.csv?base_name=Pteridophyta&count=genera). Save the resulting file the *project folder* where you are carrying out this analysis.
**Do not open it in Excel** as this program can distort some information.

```{r import diversity data downloaded}
Pteridophyta_div <- read.csv(file = "pbdb_data.csv",
                             header = T,
                             sep = ",")
```

Look at this new dataset. What are rows? What columns are provided?

There are a few columns with mysterious names. If you **really** need to know what these symbols are, you can find it in the rather technical article by [Foote (2000)](https://www.cambridge.org/core/services/aop-cambridge-core/content/view/7FE12B43106FC20C9AFC4031F72A56F9/S0094837300026890a.pdf/origination-and-extinction-components-of-taxonomic-diversity-general-problems.pdf). For this practical you are concerned with the column `X_bt`, which is the number of taxa *inferred* to range though (i.e. which existed but had not been sampled).

Make a plot of Pteridophyta genus richness, comparing sampled in bin and range through richness. Use the mid-point of each age as the time coordinate.

```{r calculate the mid-point}
calc_midpoint <- function(df) {
  for (i in 1:nrow(df)) {
    df$age[i] <- df$min_ma[i] + (df$max_ma[i] - df$min_ma[i])/2
  }
  return(df)
}
```

Apply this function to the dataset:

```{r apply "calculate the mid-point" to Pteridophyta}
Pteridophyta_div <- calc_midpoint(Pteridophyta_div)
```

Plot the results:

```{r plot comparison}
plot(x = Pteridophyta_div$age,
     y = Pteridophyta_div$sampled_in_bin+Pteridophyta_div$X_bt+Pteridophyta_div$implied_in_bin,
     col = "red",
     lwd = 2,
     type = "l",
     xlab = "Age [Ma]",
     ylab = "Genus richness")
lines(x = Pteridophyta_div$age,
     y = Pteridophyta_div$sampled_in_bin,
     col = "blue",
     lwd = 2,
     lty = 3)
legend("topright",
       legend = c("Range-through richness", "Sampled in bin richness"),
       lty = c(1,3),
       lwd = c(2,2),
       col = c("red", "blue"))
```

## 6. Calculate the genus richness for Pinophyta and Magnoliophyta

You now understand the nature of fossil occurrence data. It has gaps and the age is not always known precisely. But will it be better for larger groups of plants?

- Create a graph in which sampled in bin and range-through genus richness is shown as two different lines in the same plot. Make this plot for each of the two groups. 

- This plot uses mid-point age of each geochronological age. These ages have uneven durations: some are very short (e.g. Ludfordian), some are long (e.g. Famennian). How does this affect the richness plot?

- Which of these two groups is more evenly sampled in the database?

- Calculate the average duration across genus occurrences in both groups. If they are different, why might it be?


